---
title: 'Using JAGS within R: visiting classic BUGS Examples'
output: pdf_document
date: 'May 28, 2025'
author: 'Mauro Gasparini'
---

A lot of different software exists to approximate posterior distributions
in complex and hierarchical Bayesian models using MCMC
(Markov Chain Monte Carlo) simulation. 
In this note we focus on JAGS, which can be considered
one of the mature implementations of several BUGS versions
developed in the last 30 years.
JAGS will be invoked within R, which is a universally known
software environment for statistical analysis. The R2jags library
will serve the purpose. 
There exists an analogous interface Python to JAGS called PyJAGS.

The reason to focus on JAGS and BUGS is that there are many "classic BUGS"
examples  which give an overview of several different (bio)statistical problems
and applications. In addition, there are also several books
focusing on different applications of BUGS, not necessarily bio.

Download and install JAGS from

https://sourceforge.net/projects/mcmc-jags/files/latest/download

Download the JAGS version of the Classic Bugs files from

https://sourceforge.net/projects/mcmc-jags/files/Examples/4.x/classic-bugs.tar.gz/download

and install them in your favorite directory (YFD).

Then get into Rstudio and install, one for all, the relevant package to run JAGS from within R:

```{r}
# No need to keep install.packages() in the script once it's in your lockfile!
library(R2jags)   
```
For illustration, we run the first example, called RATS, in the first Volume
of classic BUGS examples. 

**Note on Missing Data:** To run the following code, you must have the `classic-bugs` 
directory structure mentioned above. Currently, these files (`test1.R`, data, and models) 
are not included in this repository. 

```{r, eval=FALSE}
# Update the path below to your local JAGS examples directory
# setwd("path/to/your/classic-bugs/vol1/rats")
# source("test1.R")
```

If you had run the simulation, the output object `x` would be available for exploration:
```{r, eval=FALSE}
# plot(x)
# summary(x)
```


You will find a nice description of the dataset and the Bayesian model
in the .pdf file WinBUGS_Vol1.pdf.

Many more examples are in the three volumes, and even more complete descriptions
of the examples are contained in the following book, which also contains
some theory and extensive practical suggestions.

*The BUGS Book: A Practical Introduction to Bayesian Analysis* (2013).
by David Lunn, Chris Jackson, Nicky Best, Andrew Thomas, David Spiegelhalter.
Chapman & Hall.



