---
title: Linear Models in R, the insulate case study
author: 
 - Mauro Gasparini^[Politecnico di Torino, mauro.gasparini@polito.it]
date: "MSM, April 1, 2026"
output: pdf_document
description: R markdown script to build a linear model
---

## R Markdown as notebook

R Markdown file ```*.rmd``` are notebook style readable into Rstudio,
producing ```*.pdf``` files like this one when processed using ```knit```.

## The insulate case study

The insulate database contains data about fuel consumption to
heat a certain building. Assume (a big assumption!) that
we have two random samples of weeks, before (26 weeks) and after (30 weeks)
some insulating work was done on the building:
in these days we recorded:

* quando: an indicator for before/after
* Temp = average weekly external temperature in Celsius 
* Cons: fuel consuption (in ft^3). 

We want to study a linear model explaining Cons as a function
of quando and Temp.



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# No need for hardcoded setwd() if we use relative paths from the script location
# setwd("../Data") 
```

## Importing the data from a file and exploring it

```{r}
# Path updated to point to the Data folder relative to this script
insulate=read.table("../Data/insulate_data.txt",
                    col.names=c("quando","temp","cons"))

# take a look at the variable in the database
str(insulate)        # notice quando is a factor with two levels

# focus on this data frame
attach(insulate)

# Negative relationship between temp and cons
plot(temp,cons) 

# Visualizing insulation as well, with some prettyfication
plot(temp,cons,type="n")
points(temp[quando=="prima"],cons[quando=="prima"],pch=1)
points(temp[quando=="dopo"],cons[quando=="dopo"],pch=2,col=2)
title("With and without insulation")
```

## Linear models

Linear model with one quantitative and one binary predictor.
Each predictor appear to be necessary when the other is there,
since all p-values are close to zero.
The all-or-nothing (ANOVA) test is also significant.

```{r}
regr=lm(cons~quando+temp)  ### notice nonalgebraic use of symbol +
summary(regr)
```

Here are confidence intervals for the regression coefficients
```{r}
confint(regr, level=0.95)
```

Now take a look at residuals
```{r}
res=regr$resid
plot(temp,res) # no obvious pattern appears, ok
hist(res)
```

Redo it all together.
```{r}
plot(temp,cons,type="n", xlab="temperatura", ylab="consumo")
points(temp[quando=="prima"],cons[quando=="prima"],pch=16, col=1)
points(temp[quando=="dopo"], cons[quando=="dopo"], pch=17, col=2)
title("With and without insulation, additive model (parallel lines)")
# Draw the two separate fitted regression lines
abline(a=regr$coef[1]+regr$coef[2], b=regr$coef[3], lwd=2)
abline(a=regr$coef[1], b=regr$coef[3], col=2, lwd=2)
```


We now fit a model with interaction between the quantitative and the binary predictor,
i.e. a non-additive model.

```{r}
regr2=lm(cons~quando+temp+quando*temp)  ### notice the use of *
summary(regr2)
confint(regr2, level=0.95)
```

Do a similar analysis for the model with interaction

```{r}
res2=regr2$resid
plot(temp,res2) # residuals are still ok

plot(temp,cons,type="n", xlab="temperatura", ylab="consumo")
points(temp[quando=="prima"],cons[quando=="prima"],pch=16, col=1)
points(temp[quando=="dopo"], cons[quando=="dopo"], pch=17, col=2)
title("With and without insulation, model with interaction")
abline(a=regr2$coef[1]+regr2$coef[2], b=regr2$coef[3]+regr2$coef[4], lwd=2)
abline(a=regr2$coef[1], b=regr2$coef[3], col=2, lwd=2)
```

## Several F tests

```{r}
nullo <- lm(cons~1)
solotemp <- lm(cons~temp)
additivo <- lm(cons~temp+quando)
interattivo <- lm(cons~temp*quando)

# null vs interattivo
anova(nullo, interattivo) 
summary(interattivo) # same as all-or-nothing test in summary(interattivo)

# additivo vs interattivo
anova(additivo, interattivo) 
summary(interattivo)# same as interaction test in default view

# solotemp vs interattivo
anova(solotemp, interattivo) # nested models

## detach the database when you do not use it any more
detach(insulate)
```

## Confidence and Prediction intervals


```{r}
predict.lm(additivo,
           newdata=data.frame(quando=c("prima","dopo"),temp=rep(3.2,2)),
           interval="confidence",
	   level=.99)

predict.lm(additivo,
           newdata=data.frame(quando=c("prima","dopo"),temp=rep(3.2,2)),
           interval="prediction",
	   level=.99)
```


## Cool plots of confidence and prediction bands

More modern R coding style: using tibbles instead of data frames,
using library dplyr for data manipulation and using ggplot for cool plots.


```{r}
library(dplyr)
library(ggplot2)
# build x grid
dopo_insulate <-
  insulate %>%
  dplyr::filter(quando == "dopo") %>%
  dplyr::select(-quando)

# fit a simple model
dopo_simple_lm <- lm(cons ~ temp, data = dopo_insulate)

# build a vector of new data spanning the whole temp support
new_x <- tibble(temp = seq(min(dopo_insulate$temp), max(dopo_insulate$temp), by = 0.05))

# find confidence and prediction intervals for all of the range elements
# and change the column names to make them distinguishable
new_pred <- predict(dopo_simple_lm, newdata = new_x, interval = "prediction") %>%
  as_tibble() %>%
  rename_with(~ paste(.x, "pred", sep = "_")) #
new_conf <- predict(dopo_simple_lm, newdata = new_x, interval = "confidence") %>%
  as_tibble() %>%
  rename_with(~ paste(.x, "conf", sep = "_"))

# join the two interval details
new_data <- bind_cols(new_x, new_pred, new_conf)

ggplot() +
  geom_point(aes(temp, cons), data = dopo_insulate) + # scatter plot
  geom_line(aes(temp, fit_conf), data = new_data) + # regression line
  geom_ribbon(aes(temp, ymin = lwr_pred, ymax = upr_pred, fill = "prediction"),
    data = new_data, alpha = .5
  ) + # pred intervals
  geom_ribbon(aes(temp, ymin = lwr_conf, ymax = upr_conf, fill = "confidence"),
    data = new_data, alpha = .5
  ) # conf intervals

```