---
title: Two qualitative predictors
date: "MML, April 21, 2026"
author: Mauro Gasparini^[Politecnico di Torino, mauro.gasparini@polito.it]
output: pdf_document
---

This is a pdf file "260421 Two factors.pdf"
produced by the "269421 Two factors.rmd" file when "knit" within R,
according to the notebook style of R Markdown.

## Two factors: a linear model with two qualitative predictors

We recycle a textbook example taken from
McClave JT., Benson PG. e Sincich T. (2014). 
*Statistics for Business and Economics.* 
Pearson Education Limited.   
We want to study the effect on the response (variable *distance*)
of 4 different brands of golf ball (A,B,C,D) (variable *brand*)
and the club  type (DRIVER/IRON) (variable *club*).
These two features are qualitative predictors, also called factors.
A robot player is used.

Let's build the data in *wide* format, i.e. enter the data online.
The data shows 4 replications for each pair of brand and club)

```{r}
golfwide <- read.table(header=T, text='
club    A       B       C     D
DRIVER 	226.4	238.3	240.5	219.8
DRIVER 	232.6	231.7	246.9	228.7
DRIVER 	234.0 227.7	240.3	232.9
DRIVER	220.7	237.2	244.7	237.6
IRON   	163.8	184.4	179.0	157.8
IRON   	179.4	180.6	168.0	161.8
IRON   	168.6	179.5	165.2	162.1
IRON   	173.4	186.2	156.5	160.3
')
```

To tranform from wide format to long format,
we use the library *tidyr()*.

```{r}
# No need to keep install.packages() in the script once it's in your lockfile!
library(tidyr)
golflong <- gather(golfwide, brand, distance, A:D) 
golflong
```

Let us also use the old-fashioned attach/detach dynamics
and study first the effect of **brand** only
(a linear model with one factor only)

```{r}
attach(golflong)
summary(lm(distance ~ brand))
anova(lm(distance ~ brand))
model.matrix(lm(distance~brand)) ### this is the design matrix
```

To visualize dta, we can use boxplots or more specialized graphics
for qualitative predictors (for the latter, 
we have to define the predictors explicitely as factor objects in R).
Recall that prettier graphics can always be produced using *ggplot2()*.

```{r}
boxplot(distance ~ brand) 
brand <- as.factor(brand)  
club <- as.factor(club)
plot.design(distance ~ brand)
```

## Additive and non-additive models

Now let us add the second factor *club* and build a 'small'
additive model ...

```{r}
plot.design(distance ~ brand  + club)
small <- lm(distance ~ brand  + club)
summary(small)
anova(small)
model.matrix(small)
```

... and a larger model with interaction.
An interaction plot suggests interactions may be significant,
but we have to test for it since interaction plots are
subject to sampling variation.

```{r}
#plot.design(distance ~ brand*club)
interaction.plot(brand, club, distance)
large <- lm(distance ~ brand*club)
summary(large)
model.matrix(large)
anova(large)
```

With the last *anova()* command we have build a more
traditional anova table.
We can compare the two models also with the *anova* command.

```{r}
anova(small, large)
```

Interactions are significant, meaning each combination of the two factors
tells a different story.

```{r}
fitted(large)
```


## Fitted and predicted value, with confidence and prediction intervals

The following code is an application of the concepts of fitted values,
confidence intervals, predictive intervals.


```{r}
predict.lm(large, newdata=data.frame(brand="A",club="DRIVER"))

predict.lm(large, newdata=data.frame(brand="A",club="DRIVER"),
           interval="confidence",
           level=.99)

predict.lm(large, newdata=data.frame(brand="A",club="DRIVER"),
           interval="prediction",
           level=.99)

detach(golflong)
```


